---
title: "Demo of Leaflet in R"
author: "Nick Mader"
date: "July 11, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 5, fig.height = 3, warning = FALSE, message = FALSE, error = FALSE) # root.dir = "~/GitHub/leaflet-trainr/" ... currently not sure how to reliably get Knitr to respect this identification of root.directory. See https://github.com/yihui/knitr/issues/277
library(leaflet)
library(rgdal)
library(readxl)
library(RColorBrewer)
library(shiny)
library(foreign)
```

<!-- Note to self, here are some nice references:
* for using knitr and RMarkdown -- http://kbroman.org/knitr_knutshell/pages/figs_tables.html 
* the RMarkdown cheat sheet -- https://www.rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf
* info on Knitr options -- http://yihui.name/knitr/options/
* RStudio's R Leaflet tutorial -- https://rstudio.github.io/leaflet/
-->

This document is intended to give a demonstration--though not full tutorial or demo of features--of how Leaflet can be used in R to easily generate attractive, content-rich, interactive maps.

Although Leaflet is a Javascript package, folks in R's open source community have created binding that allow for a great deal of Leaflet's functionality to be available to R users, which decreases the training and logistical challenges to working across platforms when seeking to visualize and communicate our other data work.

While I have strived to create a helpful sequence for learning about Leaflet's functionality in R, a more extensive tutorial on the topic can be found [here](https://rstudio.github.io/leaflet/).

## Minimal Coding Required

Creating an interactive map with colors, markers, zoom, etc is as easy as just a
few lines of code.

```{r first_map}
m <- leaflet() %>%
       addTiles() %>%  # Add default OpenStreetMap map tiles
       addMarkers(lng=-87.5947675, lat=41.7858686, popup="Chapin Hall")
m
```

## Baselayers
By default, Leaflet in `R` uses map tiles from Open Street Maps. However, it can
easily access and use map tiles from other sources and styles.

```{r first_map_newBase}
m %>% addProviderTiles("CartoDB.Positron")
``` 

## Displaying Data
It is easy to display data that has geographic components. The code below
sources data on CPS school High School report cards from the [City of Chicago Data Portal](https://data.cityofchicago.org/).

```{r map_schools}
# Download and prep CPS report card data
schools <- read.csv("https://data.cityofchicago.org/api/views/2m8w-izji/rows.csv",
                    stringsAsFactors = FALSE) %>%
  within({
    temp <- Latitude
    Latitude <- Longitude
    Longitude <- temp
    PolicyLevel <- factor(CPS.Performance.Policy.Level)
    rm(temp)
  }) # ...because these are accidentally backwards

# Download and merge an Excel file with enrollment data
download.file(url = "http://cps.edu/Performance/Documents/Datafiles/enrollment_20th_day_2014.xls",
              destfile = "sch_enrollment_xls.xls", method = "curl")
sch_enrollment <- read_excel("sch_enrollment_xls.xls")
sch_merge <- merge(x = schools,
                   y = sch_enrollment,
                   by.x = "School.ID",
                   by.y = "School ID")

# Generate the map
leaflet(sch_merge) %>% addTiles() %>%
  addMarkers(lng = ~Longitude, lat = ~Latitude, popup = ~Name.of.School)
```

## Symbolizing Data
<!-- can color markers or size them by adding just a bit of code 
# palette <- brewer.pal(n = length(levels(sch_merge$PolicyLevel)),
#                       name = "Set1")
-->
```{r map_symbology}
# Picking out a color palette
palette <- colorFactor(palette = "Set1",
                       domain = sch_merge$PolicyLevel)

# Changing our display from markers to circles, and adding the formatting information we want, about color, radius, outline, and opacity
sch_map_withFormats <- leaflet(sch_merge) %>% addProviderTiles("CartoDB.Positron") %>%
  addCircles(lng = ~Longitude, lat = ~Latitude, popup = ~Name.of.School,
             color = ~palette(PolicyLevel), radius = ~Totals/3, stroke = FALSE,
             opacity = 0.8, fillOpacity = 0.8)
sch_map_withFormats
```

## Formatting Display
Adding a legend:
```{r map_legend}
# Using the map that we just created, but adding a legend, giving it the information on colors, values, and title that we want.
sch_map_addedLegend <- 
  sch_map_withFormats %>%
    addLegend(position = "bottomright", pal = palette, values = ~PolicyLevel,
              title = "CPS School Policy Levels")
sch_map_addedLegend
```

## Generating Informative Pop-Ups
We can create informative pop-ups using any of the information in out data sets,
using basic HTML to create formats and web links.
```{r map_detailedPopup}
detailedPopUp <- with(sch_merge, paste(sep = "<br/>",
                                       paste0("<b>", Name.of.School, "</b> (<a href='", Website,"'>web</a>)"),
                                       paste0(Street.Address, " ", City, ", ", State, " " , ZIP.Code),
                                       paste0("Network: ", Network),
                                       paste0("Policy Level: ", PolicyLevel)))
map_wPops <- leaflet(sch_merge) %>% addProviderTiles("CartoDB.Positron") %>%
               addCircles(lng = ~Longitude, lat = ~Latitude, popup = ~detailedPopUp,
                          color = ~palette(PolicyLevel), radius = ~Totals/3, stroke = FALSE,
                          opacity = 0.8, fillOpacity = 0.8)
map_wPops
```

## Map Underlays
<!-- Add polygons for census tracts-->
Just as we can add points, we can also work with polygons. Here, we use R to
read a shape file and merge in ACS data that we will use to create data
symbolization. 

```{r map_polygonSymbology}
# First, we load the rgdal package--the R Geo-spatial Data Abstraction Library--
# which will allow us to do a variety of GIS data manipulations
library(rgdal) 

# Next, we load in a shape file representing all of the Census tracts in Cook County which already has ACS data merged in, and transform it into local coordinate reference system.
# (Note: The "acs" package allows for downloading and working with ACS data directly in R. We don't use that here for the sake of space.)
chiTracts <- readOGR(dsn = "../data/CensusTracts_acsMerge.shp",
                     layer = "CensusTracts_acsMerge",
                     encoding = "UTF-8",
                     verbose = FALSE)
chiTracts <- spTransform(chiTracts, CRS("+proj=longlat +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +no_defs"))

#Prepare colors for symbolizing our polygon, and add it to the map
quantpal_bl <- colorQuantile(palette = "Blues", domain = NULL, n = 5)

map_wPolys <- leaflet(sch_merge) %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(data = chiTracts,
              fillColor = ~quantpal_bl(chiTracts$rEd_LtHS),
              color = "#BDBDC3", weight = 0.5, fillOpacity = 0.8) %>%
  addCircles(lng = ~Longitude, lat = ~Latitude, popup = ~detailedPopUp,
             color = ~palette(PolicyLevel), radius = ~Totals/3, stroke = FALSE,
             opacity = 0.8, fillOpacity = 0.8)
map_wPolys

```

## Displaying Layers

```{r map_layerSelection}
quantpal_gr <- colorQuantile(palette = "Greens", domain = NULL, n = 5)
map_wLayers <- leaflet(sch_merge) %>%
  addProviderTiles("CartoDB.Positron") %>%
  addCircles(lng = ~Longitude, lat = ~Latitude, popup = ~detailedPopUp,
             color = ~palette(PolicyLevel), radius = ~Totals/3, stroke = FALSE,
             opacity = 0.8, fillOpacity = 0.8, group = "Schools") %>%
  addPolygons(data = chiTracts,
              fillColor = ~quantpal_bl(chiTracts$rEd_LtHS),
              color = "#BDBDC3", weight = 0.5, fillOpacity = 0.8,
              group = "% Ed < HS") %>%
  addPolygons(data = chiTracts,
              fillColor = ~quantpal_gr(chiTracts$rNonWhite),
              color = "#BDBDC3", weight = 0.5, fillOpacity = 0.8,
              group = "% Non-White") %>%
  addLayersControl(
    baseGroups = c("% Ed < HS", "% Non-White"),
    overlayGroups = c("Schools"),
    options = layersControlOptions(collapsed = FALSE))
map_wLayers
```


